---
title: "A catalog of datasets for environmental economics"
description: "A curated list of datasets useful for applied environmental economics research"
author: "Patrick Baylis"
date: 2022-05-13
format: 
  html:
    page-layout: full
categories:
  - data
---

With the help of Keith Kirchner, one of our all-star graduate students, I've started to put together a catalog of datasets relevant to environmental economics research. I teach several classes where research design is a main emphasis, and many students have found it useful to have a list of datasets that are out there as they begin thinking about potential research ideas.

This post (a work in progress) captures our initial efforts to pull together a list. The loose inclusion criteria are data that might be useful for applied environmental economics research. The imagined audience is students in one of my courses, but I'm hopeful that it will be useful for others as well. Feel free to [send me a note](mailto:patrick.baylis@ubc.ca) with other datasets that are useful.

```{r setup, include=F}
knitr::opts_chunk$set(echo = F)
library(tidyverse)
library(DT)
```

<br>

```{r,  include=F}
sample <- read.csv("env-econ-datasets.csv")
colnames(sample) = c('Source' , 'URL')
sample$URL = paste0('<a href="',sample$URL,'" target="_blank">',sample$Source,'</a>')
sample = subset(sample, select = -c(Source))
colnames(sample) = c('Source', 'Category', 'Sub-category', 'Region', 'Start Date', 'End Date', 'Description')
```

```{r}
sample %>%   datatable(extensions = 'Buttons', filter = 'top', escape = FALSE,
            options = list(dom = 'Blfrtip',
                           buttons = c('csv', 'excel', 'pdf'),
                           lengthMenu = list(c(10,25,50,-1),
                                                 c(10,25,50,"All"))))

```

<br>

### Other catalogs

- JPAL-North America's [Catalog of Administrative Datasets](https://www.povertyactionlab.org/catalog-administrative-data-sets)

