---
layout: post
title:  "Linear combinations of coefficients in R"
date:   2018-04-12
categories: 
  - coding
tags: 
  - r
  - prediction
---

Updated: 2020-12-05

One of the few features I miss from Stata is the very-intuitive `lincom` command. Most of the time I recreate that functionality with `survey::svycontrast`. But, I always forget the syntax. This code demonstrates a minimal working example. 

```{r survey, message = F, warning = F, error = F}
library(survey)

N <- 100
df <- data.frame(x1 = rnorm(N), x2 = rnorm(N))
df$y <- 1 + 3 * df$x1 - 2 * df$x2 + rnorm(N, 0, 0.5)

fit <- lm(y ~ x1 + x2, data = df)
```

Having fit the model, we can pass unnamed vector with the right number of coefficients to get our desired linear combination:

```{r result1}
svycontrast(fit, c(0, 1, 1))
```

Or a named vector with any number of coefficients (as long as the names match). Or a list of named vectors. One "gotcha" to keep in mind: I have found that the latter syntax fails for some versions of `survey` and may be OS-dependent. 

```{r result2}
svycontrast(fit, list(c("x2" = 1, "x1" = 1), c("x2" = 4, "x1" = 2)))
```

If you want an easy way to pass `svycontrast` a `data.frame` of where each row is a different linear combination (and each column a different variable), below is how you do that. This is usually if you want to produce similar behavior to, say, running `predict` on a fitted model but A) the predict call for that model doesn't return standard errors (as with `felm` or `fixest`) and/or B) you only want to linearly combine *some* of the variables (equivalent to setting all other coefficients = 0). In either case, this will work:

```{r result3}
svycontrast_df <- function(fit, newdata) {
  # Call surveyconstrast with a data frame 
  df <- newdata
  
  # Transform data.frame into a list of its row vectors
  df_list <- as.list(as.data.frame(t(newdata))) 
  df_list <- lapply(df_list, setNames, colnames(df)) # Set all character vector names inside list
  
  # Return a named list
  setNames(as.list(as.data.frame(svycontrast(fit, df_list))), c("est", "se"))
}

newdata <- data.frame(x1 = seq(-3, 3, 1))
newdata$x2 <- newdata$x^2
svycontrast_df(fit, newdata)
```


