---
title: "How to plot a specification curve"
description: "A simple tidyverse approach to creating specification curves for sensitivity analysis"
author: "Patrick Baylis"
date: 2020-02-28
format: 
  html:
    page-layout: full
categories:
  - R
  - visualization
  - econometrics
---

::: {.callout-note}
# Updates

- Updated 2025-11-24 to account for breaking package changes.
:::

Like many researchers, I often want to plot a range of coefficient estimates to figure out if the results I'm finding are robust to other sensible specification and functional form choices. This kind of estimate is called a specification curve ([Simonsohn, Simmons, and Nelson 2015](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2694998)), and I [am](https://twitter.com/ArielOrtizBobea/status/1233224906709618689) [far](https://twitter.com/hhsievertsen/status/1188780383736909825) [from](https://twitter.com/deaneckles/status/1171175422219866112) [the](https://twitter.com/icymi_r/status/1208660197763362816) [first](https://twitter.com/JoachimGassen/status/1122410367915823104) [to](https://twitter.com/Bachl/status/1200325153902780417) [do](https://twitter.com/JoachimGassen/status/1203326365715619840) [it](https://twitter.com/thaddunning/status/1146505592682504194). In fact, there are even a couple packages available: [Joachim Gassen's `rdfanalysis`](https://github.com/joachim-gassen/rdfanalysis) and [Philipp Masur's `specr`](https://github.com/masurp/specr) (I haven't used either, yet).

I wanted to roll my own, though. Since my code is fairly simple (less than 60 lines including comments and generous spacing) and uses the [`tidyverse`](https://www.tidyverse.org/), it may be helpful to other people too. 


```{r setup, message = F, warning = F, echo = T}
pacman::p_load(tidyverse, cowplot, fastDummies)
theme_set(theme_cowplot())
set.seed(42)
```

First, we create some fake estimates. This includes the point estimates, standard errors, and any variables that describe the specifications we "estimated".

```{r create-fake-data, message = F, warning = F, results='hide'}
fe <- c("Unit", "Time", "Unit + Time")
controls <- c("Basic", "Some", "Full")

estimates <- as_tibble(expand.grid(`Fixed Effects` = fe, `Controls` = controls))
estimates <- estimates %>% mutate(est = rnorm(n()), se = runif(n(), 0, 0.1))
estimates <- estimates %>% mutate(ci_l = est - 1.96 * se, ci_h = est + 1.96 * se)
```

Next, we construct a typical coefficient plot that shows how the estimates and confidence intervals change across different categories.

```{r plot-curve, message = F, warning = F, results='hide'}
spec_cols <- c("Fixed Effects", "Controls") # Could be set by user or figured out

estimates <- estimates %>%
  arrange(est) %>%
  mutate(h_order = 1:n()) # Sort on point estimates for horizontal ordering

coef_plot <- ggplot(estimates, aes(x = h_order, y = est)) +
  geom_linerange(aes(ymin = ci_l, ymax = ci_h), linewidth = 1, alpha = 0.5) +
  geom_point(fill = "white", shape = 21) +
  labs(y = "Coefficient") +
  theme(axis.title.x = element_blank(), axis.ticks.x = element_blank(), axis.line.x = element_blank(), axis.text.x = element_blank())
```

Then we visualize the specifications themselves as a series of dot plots below the coefficient plot.

```{r plot-specs, message = F, warning = F, results='hide'}
make_spec_plot <- function(category) {
  # Creates a dot plot for one specification category
  specs <- dummy_cols(estimates, select_columns = category, remove_selected_columns = T) %>%
    select(h_order, starts_with(category)) %>%
    pivot_longer(starts_with(category), names_prefix = paste0(category, "_")) %>%
    mutate(name = factor(name, levels = rev(unique(name))))

  spec_plot <- ggplot(specs, aes(x = h_order, y = name, alpha = value)) +
    geom_point() +
    scale_alpha_continuous(guide = "none") +
    theme(axis.title.x = element_blank(), axis.ticks.x = element_blank(), axis.line.x = element_blank(), axis.text.x = element_blank()) +
    theme(axis.text.y = element_text(size = 6), axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())
  spec_plot
}
spec_plots <- lapply(spec_cols, make_spec_plot)
```

Finally, we combine them for the final result:
```{r combine-plots, message = F, warning = F, results='hide'}
combined_plot <- plot_grid(
  plotlist = c(list(coef_plot), spec_plots),
  labels = c("", spec_cols), label_size = 8, label_fontface = "italic", vjust = 0.5, hjust = -0.1,
  rel_heights = c(1, 0.2, 0.2), ncol = 1, align = "v"
)

combined_plot
```

This approach is flexible and can be easily adapted to include more specification categories, different aesthetics, or other customizations. Happy plotting!
